---
title: "Data_Viz_Lesson_5_ggplot"
author: "Willy Feid"
date: "2023-10-25"
output:
  html_document: default
  pdf_document: default
  word_document: default
---

```{r message=FALSE, warning=FALSE, include=FALSE}
#Libraries
library(readr)
library(tidyverse)
library(knitr)
library(plotly)
library(car)
library(lattice)
library(MASS)
library(parcoords)
library(ggthemes)
library(shiny)
library(extrafont)
library(dplyr)
library(scales)
library(skimr)
library(lubridate)
library(extrafont)
```


# 1. Data Exploration and Cleaning

---

```{r message=FALSE, warning=FALSE, include=FALSE}
# Read the CSV file
bikeshare <- read_csv("../Data/bikesharedailydata.csv")
```

```{r message=FALSE, warning=FALSE, include=FALSE}
head(bikeshare)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
#skim function from skimr package
skim(bikeshare)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
# season and mnth have 1 NA value each
na_rows <- bikeshare[is.na(bikeshare$season) | is.na(bikeshare$mnth), ]
print(na_rows)

# The `season` NA should be replaced with 1 (for winter)
# The `mnth` NA should be replaced with 1 (for Janaury)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# get all column names
colnames(bikeshare)
```


```{r message=FALSE, warning=FALSE, include=FALSE}
# check unique values of certain columns
unique_values <- lapply(bikeshare[c("season", "yr", "mnth", "holiday", "weekday", "workingday", "weathersit")], unique)
unique_values
```

**Data Summary & Cleaning**

  - `instant`: This is an index
    - Rename to ID
  - `dteday`: date column 
    - Rename to Date
  - `season`: Values 1-4. By looking at the data, we can determine that 1: Winter, 2: Spring, 3: Summer, 4: Winter
    - Rename seasons to Winter, Spring, Summer, Fall
    - Impute NA value with 1 (season NA value has data 1/7/11)
  - 'yr': Has values 0 and 1. 0 = 2011, 1 = 2012
    - Rename to Year, and values to 2011 and 2012
  - `mnth`: Values 1-12 for month
    - Rename to Month, and values to month names.
    - Impute NA value with 1  for January (date for this row is 1/10/11)
  - `holiday`: 1= Holiday, 0 = Not Holiday
  - `weekday`: 0=Sun, 1=Mon, ...., 6=Sat
    - Re label values to names of day of the week.
  - `workingday`: 1=workday, 0=weekend or holiday.
  - `weathersit`: values of 1, 2, 3. Seem to indicate general weather (ie, clear, cloudy, etc)
  - `temp`: Temperature that has been normalized.
  - `atemp`: Could be average temp or feeling temp. Also normalized.
  - `hum`: Humidity, normalized
  - `windspeed`: Wind speed normalized.
  - `casual`, `registered`, `cnt`: casual + registered = cnt. We can infer that cnt is the total number of bike rentals for that day, and casual and registered are two types of renters.
  
**Cleaning**

  
```{r}
# Fill NA values:
bikeshare$season[is.na(bikeshare$season)] <- 1
bikeshare$mnth[is.na(bikeshare$mnth)] <- 1

# Rename columns:
bikeshare <- bikeshare %>%
  rename(
    index = instant,
    date = dteday,
    year = yr,
    month = mnth,
    humidity = hum,
    rentals = cnt
  )

# Rename Values for season, year, and weekday
bikeshare <- bikeshare %>%
  mutate(
    season = case_when(
      season == 1 ~ "Winter",
      season == 2 ~ "Spring",
      season == 3 ~ "Summer",
      season == 4 ~ "Fall"),
    year = case_when(
      year == 0 ~ "2011",
      year == 1 ~ "2012"),
    weekday = case_when(
      weekday == 0 ~ "Sunday",
      weekday == 1 ~ "Monday",
      weekday == 2 ~ "Tuesday",
      weekday == 3 ~ "Wednesday",
      weekday == 4 ~ "Thursday",
      weekday == 5 ~ "Friday",
      weekday == 6 ~ "Saturday"),
    month = month(month, label = TRUE)
  )

# Reorder the factor levels of the date column
bikeshare$date <- factor(bikeshare$date, levels = unique(bikeshare$date))

# Subset for 2011
bikeshare_2011 <- bikeshare %>% 
  filter(year == "2011")

# Subset for 2012
bikeshare_2012 <- bikeshare %>% 
  filter(year == "2012")

head(bikeshare)
```



# 2. Visualizations

---

```{r}
# Defining chart attributes

chartcolor <- "#4E79A7"
chartcolor2 <- "#053661"
chartcolor_hunter <- "#589d62"
chartcolor_hunter2 <- "#467c4e"
chartcolor_gray <- "#5a5a5a"
caption <- "Source: Bikeshare Dataset (2011 - 2012) | Willy Feid"
caption_2011 <- "Source: Bikeshare Dataset (2011) | Willy Feid"
caption_2012 <- "Source: Bikeshare Dataset (2012) | Willy Feid"
mychartattributes <-   theme_minimal() +
  theme(text=element_text(family = ".New York")) +
  theme(panel.border = element_blank(), 
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.line = element_line(color = "gray"),
        axis.line.y = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        plot.title.position = "plot",
        plot.caption = element_text(size = 8, color=chartcolor_gray, ),
        plot.subtitle = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10, color = chartcolor_gray),
        axis.title.y = element_text(size = 10, color = chartcolor_gray))
```


### Density Plot

---

```{r}
caption = "Source: Bikeshare Dataset (2011 - 2012) | Willy Feid"
ggplot(bikeshare, aes(x=rentals)) +
  geom_density(
    fill = chartcolor, color = chartcolor) + #add fill color
  labs(title = "Density Plot of Daily Bike Rentals",
       subtitle = "Daily bike rental fall around the mean (4,504) with notable smaller peaks above and below the mean.",
       caption = caption,
       x = "Rentals",
       y = "Density") +
  mychartattributes +
  geom_vline(xintercept = round(mean(bikeshare$rentals),2), linewidth=0.5, color = "white", linetype = "dashed")
```


### Histogram

---

```{r}
ggplot(bikeshare, aes(x=rentals)) +
  geom_histogram(binwidth = 100, # set thickness of bins
                 bins = 100,
                 fill = chartcolor) +
  labs(title = "Histogram of Daily Bike Rentals from 2011 to 2012",
       subtitle = "Daily bike rental fall around the mean but there are peaks of very busy days, and \nvery slow days (could be weather or seasonal conditions).",
       caption = caption,
       x = "Rentals",
       y = "Frequency") +
  mychartattributes +
  geom_vline(xintercept = round(mean(bikeshare$rentals),2), linewidth=0.5, color = "white", linetype = "dashed") +
  scale_x_continuous(labels = comma) +
  theme(axis.title.x = element_text(vjust = 3, hjust = 1),
        axis.title.y = element_text(angle = 90, vjust = 0, hjust = 0.5))
```


### Bar Chart

---

**Bike Rentals by Month between 2011 & 2012**
```{r}
# Getting the sum of rentals for each month (2011 + 2012)
monthly_rentals <- bikeshare %>%
  group_by(month) %>%
  summarise(total_rentals = sum(rentals))
# Plot
ggplot(monthly_rentals, aes(x=month, y=total_rentals)) +
  geom_bar(stat="identity", fill = chartcolor) +
  labs(title = "Bike Rentals by Month between 2011 & 2012",
       subtitle = "Bike rentals show a seasonality trends.Rentals are most popular in late summer and early fall, with \nclear drop offs in colder winter months.",
       caption = caption,
       x = " ",
       y = " ") +
  mychartattributes +
  scale_y_continuous(labels = comma)
```



**Rentals per Month by Year**
```{r}
ggplot(bikeshare) +
  geom_bar(aes(x=rentals, y=month), stat="identity", fill = chartcolor) +
  facet_grid(.~year) +
  labs(title = "Bike Rentals by Month in 2011 and 2012",
       subtitle = "Seasonality factors were consistent in both years, but overall rentals increased in 2012.",
       caption = caption,
       x = " ",
       y = NULL) +
  theme_minimal() +
  theme(
    text=element_text(family = ".New York"),
    plot.title.position = "plot",
    plot.subtitle = element_text(size=10),
    strip.text = element_text(size=12, hjust = 0.04),
    plot.caption = element_text(hjust = 1, vjust = 1, size = 8, color="#5a5a5a"),
    panel.grid.major = element_line('white', linewidth = 0.5),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.ontop = TRUE
  ) +
  scale_x_continuous(labels = comma)

```


### Box Plot

---

```{r}
ggplot(bikeshare_2012, aes(x=month, y=rentals)) +
  geom_boxplot(color = chartcolor, fill = "#cccccc") +
  labs(title = "Monhtly Distributions of Bike Rentals in 2012",
       subtitle = "Higher variability in less popular monthly and more consistency in popular months.",
       caption = caption_2012,
       x = " ",
       y = "Count") +
  mychartattributes +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.9, hjust = 1)) +
  scale_y_continuous(labels = comma)
  
```

### Line Chart

---

```{r}
# for x-axis ticks
first_date <- "1/1/12"
mid_date <- "7/2/12"
last_date <- "12/31/12"
bikeshare_2012$holiday <- factor(bikeshare_2012$holiday, levels = c(0, 1))
#plot
ggplot(bikeshare_2012, aes(x=date, y=rentals, group=1)) +
  geom_line(color = chartcolor) +
  geom_point(aes(color=factor(holiday))) +
  labs(title = "Daily Bike Rentals in 2012",
       subtitle = "Holidays are not outliers but instead stay along the general trend.",
       caption = caption_2012,
       x = " ",
       y = "Count",
       color = NULL) +
  mychartattributes +
  theme(axis.title.y = element_text(angle = 0, vjust = 0.9, hjust = 1)) +
  scale_x_discrete(breaks = c(first_date, mid_date, last_date), expand = c(0.05, 0.05)) +
  scale_color_manual(values = c("0" = chartcolor, "1" = "#ff8989"),
                     breaks = "1",
                     labels = "Holiday") +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = comma)
```


### Scatter Plot

---

```{r}
ggplot(bikeshare) +
  geom_point(aes(x = temp, y = casual, color = "Casual")) +
  geom_point(aes(x = temp, y = registered, color = "Registered")) +
  labs(title = "Correlation of Temperature and Number of Rentals",
       subtitle = "Temperature has a greater effect on casual users than registered users. Temperature is normalized.",
       caption = caption,
       x = "Temp",
       y = NULL,
       color = NULL) +  # Optional: Remove the legend title
  mychartattributes +
  theme(
        axis.title.x = element_text(vjust = 0, hjust = 1),
        legend.position = "bottom") +
  scale_y_continuous(labels = comma) +
  scale_color_manual(values = c("Casual" = chartcolor_hunter, "Registered" = chartcolor))

```


### Stacked Area

---

**Area graph of daily rentals in 2012 by user type**
```{r}

# Reshape the data to long format
bikeshare_2012_long <- bikeshare_2012 %>%
  dplyr::select(date, casual, registered) %>%
  pivot_longer(cols = c(casual, registered), names_to = "type", values_to = "count")

# Plot the stacked area chart
ggplot(bikeshare_2012_long, aes(x = date, y = count, fill = type, group = type)) +
  geom_area(alpha = 0.6, color = NA) +
  labs(title = "Daily Bike Rentals in 2012",
       subtitle = "Registered users make up the majority of rentals, and are responsible for almost all bike rentals in the colder months.",
       caption = caption_2012,
       x = " ",
       y = " ") +
  mychartattributes +
  scale_x_discrete(breaks = c(first_date, mid_date, last_date), expand = c(0.05, 0.05)) +
  scale_fill_manual(values = c(casual = chartcolor_hunter, registered = chartcolor2)) +
  guides(fill = guide_legend(title = NULL)) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = comma)

```




```{r}
rmarkdown::render("bikeshare_visualization.Rmd", output_format = "github_document")
```

